Classless Java

### 2015-11-18 paper/sections/Formalization.tex
Formalization partly revised

\bruno{Some notes for discussion regarding formalization:
\begin{itemize}
\item Order of premisses: It may be worthwhile to swap the order of
some premisses, so that it follows the order of the explanations.
\item T-Obj: small fixes on 3rd line. Suggestion, create two
auxiliary functions: $sigvalid(C,\overline{mh})$ and $alldefined(C,\overline{mh})$
\item T-INTF: $C_0$ is $I$. Type of this was $C$, modified it to $C_0$.
In the last line, should it be $dom(\overline{meth})$ or just $\overline{meth}$?

\item tops and shadow: tops seems redundant;
we use tops to collect classes, so just that in the next step we iterate
over those classes to collect methods that satisfy some property.
I think we can merge those 2 steps.

\end{itemize}
}

\footnote{Future work: updating multiple fields in one method call,
  \texttt{with(T v)}}

\begin{figure}[h]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of ClassLess Java}
\label{Grammar}
\end{figure}

\bruno{Use space better in this figure. }

${}_{}$\\
\begin{figure}[h]
$
\!\!\!\!\!\!\!\!\!\!
\begin{array}{l}

%% T-Invk
\inferrule[(T-Invk)]{
 \Gamma \vdash \e \in \C \\\\
\forall i\in 1..n\\   \Gamma \vdash \e_i \in \_<:\C_i \\\\
  \textsf{mtype}(\m,\C) \!=\! \overline{\C} \!\!\to\! \C'
%\textsf{mmodifier}(\m,\C) \neq \textbf{static}
 }{
 \Gamma \vdash \e\QM.\m\QM(\overline{\e}\QM) \in \C' }
\quad\quad

@@

%%T-Var
\\[5ex]
\inferrule[(T-Var)]{
\Gamma(\x)=\C
}{
\Gamma \vdash \x \in\C}
\quad\quad\quad

%%T-Obj
\inferrule[(T-Obj)]{
\forall i\in 1..n\
%\Gamma_i
\Gamma,\f_1{:}\T_1,\ldots,\f_k{:}\T_k,\,\QM{this}{:}\C,\Gamma^{\mh_i}
\vdash\e_i\in \_\subtype\C^{\mh_i}\\\\
\forall i\in 1..n\ \mh_i\subtype\mBody(\m^{\mh_i},\C)\\\\
\forall i\in 1..k\ \Gamma(\x_i)\subtype\T_i\\\\
\forall\m\mbox{ such that }
\mBody(\m,\C)=\mh\QM; \exists i\in 1..n\ \m^{\mh_i}=\m
%\forall i\in 1\ldots n\ \Gamma_i=\Gamma,\f_1{:}\T_1,\ldots,\f_k{:}\T_k,\,\QM{this}{:}\C,\Gamma^{\mh_i}
}{
\Gamma \vdash\QM{new}\ \C\oR\cR\oC\T_1\ \f_1\QM=\x_1\QM;\ldots\T_k\ \f_k\QM=\x_k\QM;\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
\in\C
}
\\[5ex]

%%T-update
\quad
\inferrule[(T-update)]{
\Gamma \vdash \e\in\_<:\Gamma(\x)\\\\
\Gamma \vdash \e'\in\C
}{
\Gamma \vdash \x\QM=\e\QM;\e'\in\C }
\quad\quad\quad\quad\quad

%%T-Intf
 \inferrule[(T-Intf)]{
IT(\C) = \ann\ \QM{interface}\ \C\ \QM{extends}\ \overline{I} \
\oC\methods\ \cC\\\\
 \forall \QM{default}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \ \Gamma^{\mh},\,\QM{this}{:}\C\vdash\e\in \_\subtype\C^{\mh} \\\\
 \forall \QM{static}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \ \Gamma^{\mh}\vdash\e\in \_\subtype\C^{\mh} \\\\
\dom(\C)=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)
 }{
\C \text{ OK}
}
\end{array}$
\caption{CJ Typing}
\label{ET}
\end{figure}
\subsection{Grammar and typing rules}${}_{}$\\

In Figure~\ref{Grammar} we show the syntax of ClassLess Java.%
\footnote{To be compatible with java, the concrete syntax for an interface declaration with empty supertype list $\C_1\ldots\C_n$ would also omit the \Q@extends@ keyword.}

We formalize a minimal version of Java 8, focusing on Interfaces, default methods and object creation literals.
As you can see, we have no syntax for classes.
Expressions are the conventional variable and method call$\x$ and$\e\QM.\m\QM(\es)$, then we have conventional static method call
$\C\QM.\m\QM(\es)$; for simplicity we do not consider the degenerate case of calling a static method over the this receiver as it considered a bad/confusing programming practice.
Next we have the more interesting super call $\C\QM{.super.}\m\QM(\es)$, whose semantic is to call the (non static) method $\m$ over the $\this$ receiver, but statically dispatching to the version of the method defined in the interface $\C$.
We model field updating as $\x\QM=\e\QM;\e'$, where we expect $\x$ to be a field name. After updating the field $\x$ using the value of $\e$, the expression $\e'$ is executed. To blend the statement based nature of Java and the expression based nature of our language, we consider a method body of the form \Q@return@ $\x\QM=\e\QM;\e'$
to represent  $\x\QM=\e\QM;\QM{return}\ \e'$ in Java.
Finally, we consider the object initialization expression from an interface $\C$, where (for simplicity) all the fields are initialized with a variable present in scope.
Note how our language is exactly a subset of Java 8.
We allows the only annotation \mixin since is the only one interesting in this article.
We consider an globally present Interface Table (\metaVar{IT}) mapping from interface names $\C$ to interface declarations \metaVar{I}.

The environment $\Gamma$ is a mapping from variables to types.
As tradition, we allows functional notation for $\Gamma$.
However, to help us defining auxiliary functions,
we also allows functional notation set of methods $\methods$, using the method name $\m$ as a key.
That is, we define $\methods(\m)=\method$ iff there is a unique $\method\in\methods$ whose name is $\m$.
For convenience, we define $\methods(\m)=\none$ otherwise;
moreover $\m\in\dom(\methods)\ \miff\ \methods(\m)=\method$.

 Typing statement $\Gamma \vdash \e\in\C$ reads ``in the
environment $\Gamma$, expression $\e$ has type $\C$.''
As a shortcut, we write $\Gamma \vdash \e\in\C<:\C'$
instead of $\Gamma \vdash \e\in\C$ and $\C<:\C'$.
 From the interface table, we can read off the subtype relation between interfaces. The subtype relation is given by the \textbf{extends} clauses in the interfaces. We omit the definition of the usual, traditional subtyping relation between interfaces.%
\footnote{
Notice how there are no classes, thus there is no subclassing.
We believe that this approach may scratch an old itching point in the long struggle of subtyping versus subclassing:
According to some authors, from a software engineering perspective, interfaces are just a kind of classes. Others consider more opportune to  consider interfaces are pure types. In this vision our language would have no subclassing. We do not know how to conciliate those two viewpoints and ClassLess Java design. We do not have Classes purely in the Java sense.
}

In Figure~\ref{ET} we show the typing rules.
We use the following auxiliary notation:
$\Gamma^\mh$ trivially extract the environment from a method header, collect the types and the names of the method parameters.
$\m^\mh$ and $\C^\mh$ extracts the method name and the return type from a method header.
$\mBody(\m,\C)$ return the full method declaration as seen by $\C$, that is the method $\m$ can be declared in $\C$ or inherited from another interface.
$\mBody(\m,\C)$ will be formally defined later.
$\textsf{mtype}(\m,\C)$ and
$\textsf{mtypeS}(\m,\C)$ return the type signature from a method (using $\mBody(\m,\C)$ internally).
$\textsf{mtype}(\m,\C)$ is defined only for non static methods, while
$\textsf{mtypeS}(\m,\C)$ only on static ones.

We use $\dom(\C)$ to denote the set of methods that are defined for type $\C$, that is: $\m\in\dom(\C)\ \miff \ \mBody(\m,\C)=\method$.

We discuss first the most interesting rules, that is
\rn{(t-Obj)} and  \rn{(t-Intf)} \marco{Bruno, you can explain t-obj, if you can make it understandable, I go for t-inft}

Rule \rn{(t-Obj)}....

Rule \rn{(t-intf)} check that an interface $\C_0$ is correctly typed.
First we check that the body of all the default and static methods are well typed.
Then we check that
$\dom(\C_0)$ is the same of $\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)$.
This is not a trivial check, since $\dom(\C)$ is defined using $\mBody$, that is undefined in many cases: notably if a method $\method\in\methods$ is not compatible with some method in
$\dom(\C_1)\ldots\dom(\C_n)$ or if any method in both $\dom(\C_i)$ and $\dom(\C_j)$ ($i,j\in 1..n$)  is conflicting.

\marco{if we have space we can discuss the other rules too, but is not mandatory}

>>>>

This section presents a formalization of ClassLess Java: a minimal
FeatherweightJava-like calculus which models the essence of Java 
interfaces with default methods. This calculus also supports a minimal
extension to Java to account for the \mixin annotation for object
interfaces. The syntax, typing rules and method lookup mechanisms 
of CJ are modelled. Finally, using our formalization, we discuss
examples of real Java programs that seem to be incoherent\bruno{revise
later}.

\begin{figure}[t]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of ClassLess Java}
\label{Grammar}
\end{figure}

\subsection{Syntax and Typing}\marco{Future work: updating multiple fields in one method call,
  \texttt{with(T v)}}

\paragraph{Syntax}
Figure~\ref{Grammar} shows the syntax of ClassLess Java.  \footnote{To
  be compatible with java, the concrete syntax for an interface
  declaration with empty supertype list $\C_1\ldots\C_n$ would also
  omit the \Q@extends@ keyword.}  The syntax formalizes a minimal
version of Java 8, focusing on interfaces, default methods and object
creation literals.  There is no syntax for classes.  Expressions
consist of conventional constructs such as variables ($\x$), method
calls ($\e\QM.\m\QM(\es)$) and static method calls
($\C\QM.\m\QM(\es)$). For simplicity the degenerate case of calling a
static method over the $\this$ receiver is not considered.  A more
interesting type of expressions are super calls
($\C\QM{.super.}\m\QM(\es)$), whose semantic is to call the (non
static) method $\m$ over the $\this$ receiver, but statically
dispatching to the version of the method defined in the interface
$\C$. A simple form of field updates ($\x\QM=\e\QM;\e'$) is also
modelled. In the syntax of field updates $\x$ is expected to be a
field name. After updating the field $\x$ using the value of $\e$, the
expression $\e'$ is executed. To blend the statement based nature of
Java and the expression based nature of our language, we consider a
method body of the form \Q@return@ $\x\QM=\e\QM;\e'$ to represent
$\x\QM=\e\QM;\QM{return}\ \e'$ in Java.  Finally, there is an object
initialization expression from an interface $\C$, where (for
simplicity) all the fields are initialized with a variable present in
scope. Note how our language is a subset of Java 8.  The single
non-Java 8 piece of syntax is the \mixin annotation, which is the only
one interesting piece of syntax in this article.  Following standard
practise, we consider a global Interface Table (\metaVar{IT}) mapping
from interface names $\C$ to interface declarations \metaVar{I}.

The environment $\Gamma$ is a mapping from variables to types.  As
usual, we allow a functional notation for $\Gamma$ to do variable
lookup.  Moreover, to help us defining auxiliary functions, a functional
notation is also allowed for a set of methods $\methods$, using the
method name $\m$ as a key.  That is, we define $\methods(\m)=\method$
iff there is a unique $\method\in\methods$ whose name is $\m$.  For
convenience, we define $\methods(\m)=\none$ otherwise; moreover
$\m\in\dom(\methods)\ \miff\ \methods(\m)=\method$.

\paragraph{Typing} 
Typing statement $\Gamma \vdash \e\in\C$ reads ``in the environment
$\Gamma$, expression $\e$ has type $\C$.''. 
Before discussing the typing rules we discuss some of the used notation.
As a shortcut, we write
$\Gamma \vdash \e\in\C<:\C'$ instead of $\Gamma \vdash \e\in\C$ and
$\C<:\C'$.  From the interface table, we can read off the subtype
relation between interfaces. The subtype relation is given by the
\textbf{extends} clauses in the interfaces. We omit the definition of
the usual\bruno{where can it be found? add reference here?}, traditional subtyping relation between interfaces.%
\footnote{ Notice how there are no classes, thus there is no
  subclassing.  We believe that this approach may scratch an old
  itching point in the long struggle of subtyping versus subclassing:
  According to some authors, from a software engineering perspective,
  interfaces are just a kind of classes. Others consider more
  opportune to consider interfaces are pure types. In this vision our
  language would have no subclassing. We do not know how to conciliate
  those two viewpoints and ClassLess Java design. We do not have
  Classes purely in the Java sense.}\bruno{is a footnote here the
  appropriate place to have this discussion?}
The auxiliary notation $\Gamma^\mh$ trivially
extracts the environment from a method header, by collecting the all types
and names of the method parameters.  The
notation $\m^\mh$ and $\C^\mh$ denotes, respectivelly, extracting the
method name and the return type from a method header. $\mBody(\m,\C)$,
defined in Section~\ref{},
returns the full method declaration as seen by $\C$, that is the
method $\m$ can be declared in $\C$ or inherited from another
interface. 
$\textsf{mtype}(\m,\C)$ and $\textsf{mtypeS}(\m,\C)$ return the type
signature from a method (using $\mBody(\m,\C)$ internally).
$\textsf{mtype}(\m,\C)$ is defined only for non static methods, while
$\textsf{mtypeS}(\m,\C)$ only on static ones. We use $\dom(\C)$ to
denote the set of methods that are defined for type $\C$, that is:
$\m\in\dom(\C)\ \miff \ \mBody(\m,\C)=\method$.

\begin{figure}[t]
$
\begin{array}{l}

%% T-Invk
\inferrule[(T-Invk)]{
 \Gamma \vdash \e \in \C \\\\
\forall i\in 1..n\ \ \Gamma \vdash \e_i \in \_<:\C_i \\\\
  \textsf{mtype}(\m,\C) \!=\! \overline{\C} \!\!\to\! \C'
%\textsf{mmodifier}(\m,\C) \neq \textbf{static}
 }{
 \Gamma \vdash \e\QM.\m\QM(\overline{\e}\QM) \in \C' }
\quad\quad

@@

%%T-Var
\\[5ex]
\inferrule[(T-Var)]{
\Gamma(\x)=\C
}{
\Gamma \vdash \x \in\C}
\quad\quad

%%T-Obj
\inferrule[(T-Obj)]{
\forall i\in 1..k\ \ \Gamma(\x_i)\subtype\T_i\\\\
\forall i\in 1..n\ \ 
%\Gamma_i
\Gamma,\f_1{:}\T_1,\ldots,\f_k{:}\T_k,\,\QM{this}{:}\C,\Gamma^{\mh_i}
\vdash\e_i\in \_\subtype\C^{\mh_i}\\\\
\sigvalid(\mh_1\ldots\mh_n,I)\quad\quad\quad\quad
%\forall i\in 1..n\ \mh_i\subtype\mBody(\m^{\mh_i},\C)\\\\
\alldefined(\mh_1\ldots\mh_n,I)
%\forall\m\mbox{ such that }
%\mBody(\m,\C)=\mh\QM; \exists i\in 1..n\ \m^{\mh_i}=\m
%\forall i\in 1\ldots n\ \Gamma_i=\Gamma,\f_1{:}\T_1,\ldots,\f_k{:}\T_k,\,\QM{this}{:}\C,\Gamma^{\mh_i}
}{
\Gamma \vdash\QM{new}\ \C\oR\cR\oC\T_1\ \f_1\QM=\x_1\QM;\ldots\T_k\ \f_k\QM=\x_k\QM;\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
\in\C
}
\\[5ex]

%%T-update
\quad
\inferrule[(T-update)]{
\Gamma \vdash \e\in\_<:\Gamma(\x)\\\\
\Gamma \vdash \e'\in\C
}{
\Gamma \vdash \x\QM=\e\QM;\e'\in\C }
\quad\quad\quad

%%T-Intf
 \inferrule[(T-Intf)]{
IT(\C) = \ann\ \QM{interface}\ \C\ \QM{extends}\ \overline{I} \
\oC\methods\ \cC\\\\
 \forall \QM{default}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \ \Gamma^{\mh},\,\QM{this}{:}\C\vdash\e\in \_\subtype\C^{\mh} \\\\
 \forall \QM{static}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \ \Gamma^{\mh}\vdash\e\in \_\subtype\C^{\mh} \\\\
\dom(\C)=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)
 }{
\C \text{ OK}
}
\end{array}$
\caption{CJ Typing}
\label{ET}
\end{figure}

In Figure~\ref{ET} we show the typing rules.  We discuss first the
most interesting rules, that is \rn{t-Obj} and \rn{t-Intf}. Rule
\rn{t-Obj} is the most complex typing rule. Firstly, we need to
ensure that all field initializations are type correct, by looking up the type of
each variable assigned to a field in the typing environment and verifying that such type is a
subtype of the field type. Secondly, we check that all method bodies are
well-typed. To do this the enviroment used to check the method body
needs to be extended appropriately: we add all fields and their types;
add $\this:I$; and add the arguments (and types) of the respective
method. Finally, we need to check that all method headers are valid
with respect to the superinterfaces of $I$; and we also need to check
that all abstract and conflicted methods (that is methods, that need
to be explicitly overriden) in the superinterfaces have been
implemented. The definitions $\sigvalid$ and $\alldefined$ take care
of this: \\

$
\begin{array}{lll}
\sigvalid(\mh_1\ldots\mh_n,I) & = &
\forall i\in 1..n\ \ \mh_i\subtype\mBody(\m^{\mh_i},\C)\\

\alldefined(\mh_1\ldots\mh_n,I) & = &
\forall\m\mbox{ such that } \mBody(\m,\C)=\mh\QM; \exists i\in 1..n\ \m^{\mh_i}=\m
\end{array}
$\bruno{Marco, you probably want to double-check the explanation. Any
  more explanation needed?}
\bruno{Also, according to the text above, mbody returns the full
  method body, so in sigvalid, shouldn't we use mtype instead? The
  same for alldefined, should it be mtype? Still not very confident
  about alldefined. }

The rule \rn{(t-intf)} checks that an interface $\C$ is correctly
typed.  First we check that the body of all the default and static
methods are well typed.  Then we check that $\dom(\C)$ is the same of
$\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)$.  This is not a
trivial check, since $\dom(\C)$ is defined using $\mBody$, that is
undefined in many cases: notably if a method $\method\in\methods$ is
not compatible with some method in $\dom(\C_1)\ldots\dom(\C_n)$ or if
any method in both $\dom(\C_i)$ and $\dom(\C_j)$ ($i,j\in 1..n$) is
conflicting.
\marco{if we have space we can discuss the other rules too, but is not mandatory}
